---
title: "Protected Species Interactions — Hawaii Longline Fisheries"
subtitle: "Automated SAFE Report Section (Pelagic FEP §3.3.2) and Decision-Support Feed"
author: "SAFE Automation Working Group"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
    theme: cosmo
    code-fold: true
  docx:
    toc: true
    reference-doc: safe_reference.docx   # optional; remove if not present
params:
  use_mock: true                      # TRUE = SQLite test DB; FALSE = LOTUS via odbc
  sqlite_path: "../data/ps_monitoring_test.sqlite"
  current_year: 2025
  alpha: 0.05                         # Bayesian gate significance
  robust_threshold: 2                 # MAD-z gate
  percentile_threshold: 90            # rank-percentile gate
  use_llm: false                      # TRUE = call Gemini for narrative drafts
  llm_model: "gemini-2.5-pro"
  llm_endpoint: "https://generativelanguage.googleapis.com/v1beta"  # replace with agency-approved (e.g. Vertex AI) endpoint
  export_json: "../data/pipeline_output.json"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
suppressPackageStartupMessages({
  library(DBI); library(dplyr); library(tidyr); library(purrr); library(gt); library(ggplot2); library(jsonlite)
})
source("R/ps_functions.R")
CUR   <- params$current_year
ALPHA <- params$alpha
con   <- ps_connect(use_mock = params$use_mock, sqlite_path = params$sqlite_path)
on.exit(DBI::dbDisconnect(con), add = TRUE)

species <- ps_read(con, "species", params$use_mock)
ann     <- ps_read(con, "interactions", params$use_mock)
effort  <- ps_read(con, "effort", params$use_mock)
spatial <- ps_read(con, "spatial", params$use_mock)
ocean   <- ps_read(con, "ocean", params$use_mock)
its_ref <- ps_read(con, "its", params$use_mock)
```

::: {.callout-note collapse="true"}
## Data provenance and confidentiality
This document reads **aggregated** tables only (species × year, species × year × stratum). In production
these are LOTUS views maintained by PIFSC; in test mode (`params$use_mock = TRUE`) they come from
`ps_monitoring_test.sqlite`, a synthetic database calibrated to `ps_monitoring_tables_AI.xlsx`.
No set-level or vessel-level PIROP observer records are read, cached, or sent to any external service.
Mode for this render: **`r if (params$use_mock) "TEST (SQLite)" else "PRODUCTION (LOTUS)"`**.
:::

## 1. Interactions summary (Table 2)

```{r table2}
eff_by_year <- effort |> select(year, observed_sets, coverage_rate)

per_species <- ann |>
  left_join(eff_by_year, by = "year") |>
  mutate(rate_per_set = interactions / observed_sets) |>
  group_by(species_code) |>
  group_map(~ {
    d <- arrange(.x, year); h <- filter(d, year < CUR); c <- filter(d, year == CUR)
    pr <- gamma_poisson_prior(h$interactions)
    p_tail <- nb_tail_prob(pr[["shape"]], pr[["rate"]], c$interactions)
    qs <- nb_quantiles(pr[["shape"]], pr[["rate"]])
    ra <- mad_z(h$interactions, c$interactions); pct <- percentile_rank(h$interactions, c$interactions)
    gate <- or_gate(ra, pct, p_tail, ALPHA, params$robust_threshold, params$percentile_threshold)
    ts <- theil_sen(h$year, h$interactions)
    tibble(species_code = .y$species_code, year = CUR, current = c$interactions,
           mean = mean(h$interactions), median = median(h$interactions),
           p10 = quantile(h$interactions, .1), p90 = quantile(h$interactions, .9),
           percentile = pct, robust_anomaly = ra, n_years = nrow(h), trend_slope = ts,
           trend = ifelse(ts > 0, "Increasing", ifelse(ts < 0, "Decreasing", "Flat")),
           flag = gate$flag, gate_robust = gate$gates[["robust"]], gate_percentile = gate$gates[["percentile"]],
           gate_bayes = gate$gates[["bayes"]],
           prior_shape = pr[["shape"]], prior_rate = pr[["rate"]], p_tail = p_tail,
           pred_q05 = qs[1], pred_median = qs[2], pred_q95 = qs[3],
           rate_cur = c$rate_per_set, rate_hist = list(h$rate_per_set))
  }) |> bind_rows() |> left_join(select(species, species_code, common, group, sector), by = "species_code")

table2 <- per_species |>
  select(Species = common, Current = current, Mean = mean, Median = median, P10 = p10, P90 = p90,
         Percentile = percentile, `Robust Anomaly` = robust_anomaly, `N Years` = n_years,
         `Trend Slope` = trend_slope, Trend = trend, Flag = flag)

table2 |> gt() |>
  tab_header(title = md("**Table 2.** Observed protected-species interactions, current year vs. historical distribution")) |>
  fmt_number(columns = c(Mean, P10, P90, `Trend Slope`), decimals = 1) |>
  fmt_number(columns = c(Percentile), decimals = 0) |>
  fmt_number(columns = `Robust Anomaly`, decimals = 2, force_sign = TRUE) |>
  data_color(columns = Flag, fn = function(x) case_when(x == "High" ~ "#F18F01", x == "Low" ~ "#1C7293", TRUE ~ "#F4F6F8")) |>
  tab_source_note(sprintf("Flag = OR gate: robust anomaly ≥ %s, or percentile ≥ %s, or Bayesian tail probability < %s.",
                          params$robust_threshold, params$percentile_threshold, ALPHA))
```

## 2. Interaction rate, trend and spatial shift (Table 2b)

```{r table2b}
table2b <- per_species |>
  rowwise() |>
  mutate(tw = ifelse(group == "Sea turtle", 10, 5)) |>
  ungroup() |>
  mutate(trend_fit = map2(species_code, tw, ~ {
    d <- ann |> filter(species_code == .x, year >= CUR - .y + 1) |> left_join(eff_by_year, by = "year")
    fit_trend(d$year, d$interactions, d$observed_sets)
  })) |>
  mutate(annual_pct_change = map_dbl(trend_fit, "annual_pct_change"),
         trend_model = map_chr(trend_fit, "model"), trend_model_flag = map_chr(trend_fit, "model_flag"),
         rate_robust_anomaly = map2_dbl(rate_hist, rate_cur, ~ mad_z(.x, .y)),
         rate_percentile = map2_dbl(rate_hist, rate_cur, ~ percentile_rank(.x, .y)),
         hist_mean_rate = map_dbl(rate_hist, mean), hist_median_rate = map_dbl(rate_hist, median),
         hist_p90_rate = map_dbl(rate_hist, ~ quantile(.x, .9))) |>
  left_join(
    spatial |> group_by(species_code) |>
      summarise(hist_lon = mean(centroid_lon[year < CUR]), hist_lat = mean(centroid_lat[year < CUR]),
                cur_lon = centroid_lon[year == CUR], cur_lat = centroid_lat[year == CUR],
                pct_outside_hist90 = pct_outside_hist90[year == CUR], .groups = "drop") |>
      mutate(centroid_shift_km = haversine_km(hist_lon, hist_lat, cur_lon, cur_lat)),
    by = "species_code")

table2b |>
  transmute(Species = common, `Current Interactions` = current, `Current Rate` = rate_cur,
            `Historical Median Rate` = hist_median_rate, `Historical P90 Rate` = hist_p90_rate,
            `Rate Percentile` = rate_percentile, `Rate Robust Anomaly` = rate_robust_anomaly,
            `Annual % Change` = annual_pct_change, `Trend Model` = trend_model, `Model Flag` = trend_model_flag,
            `Trend Years` = tw, `Centroid Shift (km)` = centroid_shift_km, `% Outside Hist. 90%` = pct_outside_hist90) |>
  gt() |>
  tab_header(title = md("**Table 2b.** Interaction rate (per observed set), trend model and centroid shift")) |>
  fmt_number(columns = c(`Current Rate`, `Historical Median Rate`, `Historical P90 Rate`), decimals = 4) |>
  fmt_number(columns = c(`Rate Robust Anomaly`, `Annual % Change`, `Centroid Shift (km)`, `% Outside Hist. 90%`), decimals = 1)
```

## 3. Ocean and habitat context (Table 2c)

```{r table2c}
table2c <- ocean |>
  group_by(species_code, variable) |>
  summarise(
    current_interaction_env = interaction_env[year == CUR],
    hist_interaction_median = median(interaction_env[year < CUR]),
    current_fleet_env = fleet_env[year == CUR],
    current_selection_anomaly = current_interaction_env - current_fleet_env,
    hist_selection_median = median((interaction_env - fleet_env)[year < CUR]),
    interaction_percentile = percentile_rank(interaction_env[year < CUR], current_interaction_env),
    selection_percentile = percentile_rank((interaction_env - fleet_env)[year < CUR], current_selection_anomaly),
    historical_years = sum(year < CUR), .groups = "drop") |>
  mutate(absolute_anomaly = current_interaction_env - hist_interaction_median,
         selection_shift = current_selection_anomaly - hist_selection_median) |>
  left_join(select(species, species_code, common), by = "species_code")

table2c |>
  select(Species = common, Variable = variable, `Current Interaction Env.` = current_interaction_env,
         `Historical Median` = hist_interaction_median, `Absolute Anomaly` = absolute_anomaly,
         `Current Fleet Env.` = current_fleet_env, `Selection Anomaly` = current_selection_anomaly,
         `Selection Shift` = selection_shift, `Interaction %ile` = interaction_percentile,
         `Selection %ile` = selection_percentile) |>
  group_by(Species) |> gt() |>
  tab_header(title = md("**Table 2c.** Environment at interaction sets vs. fleet-wide environment (selection anomaly)")) |>
  fmt_number(columns = 3:8, decimals = 3) |> fmt_number(columns = 9:10, decimals = 0)
```

## 4. Bayesian anomaly check

```{r bayes-fig, fig.width=8, fig.height=4.5}
per_species |>
  select(common, current, pred_q05, pred_median, pred_q95, p_tail, flag) |>
  ggplot(aes(y = reorder(common, current))) +
  geom_linerange(aes(xmin = pmax(pred_q05, 0.5), xmax = pred_q95), colour = "#1C7293", linewidth = 3, alpha = .5) +
  geom_point(aes(x = pred_median), colour = "#065A82", size = 3) +
  geom_point(aes(x = current, colour = flag), size = 4, shape = 18) +
  scale_colour_manual(values = c(High = "#F18F01", Typical = "#21295C", Low = "#5A6C7D")) +
  scale_x_log10() + labs(x = "Interactions (log scale)", y = NULL, colour = "Flag",
       title = "Current count against the Gamma-Poisson posterior predictive 90% interval") +
  theme_minimal(base_family = "Calibri")
```

```{r bayes-table}
per_species |>
  transmute(Species = common, `Prior shape` = prior_shape, `Prior rate` = prior_rate,
            `Predictive 5%` = pred_q05, `Predictive median` = pred_median, `Predictive 95%` = pred_q95,
            `P(X ≥ current)` = p_tail, Flagged = p_tail < ALPHA) |>
  gt() |> fmt_number(columns = 2:3, decimals = 3) |> fmt_number(columns = 7, decimals = 4)
```

## 5. Multi-disciplinary context (workshop review areas)

```{r context}
ctx_spatial  <- ps_read(con, "ctx_spatial", params$use_mock)
ctx_gear     <- ps_read(con, "ctx_gear", params$use_mock)
ctx_behavior <- ps_read(con, "ctx_behavior", params$use_mock)
ctx_demo     <- ps_read(con, "ctx_demo", params$use_mock)
ctx_climate  <- ps_read(con, "ctx_climate", params$use_mock)
```

### 5.1 Spatial effort dynamics
```{r}
ctx_spatial |> gt() |> tab_header(title = "Fleet effort share by stratum, baseline vs. current")
```
### 5.2 Operational and gear modifications
```{r}
ctx_gear |> gt() |> tab_header(title = "Gear configuration, adoption and interaction outcomes") |> sub_missing(missing_text = "—")
```
### 5.3 Fisher post-interaction behaviour
```{r}
ctx_behavior |> gt() |> tab_header(title = "Post-interaction movement and trip response") |> sub_missing(missing_text = "—")
```
### 5.4 Oceanographic and climate drivers
```{r}
ctx_climate |> gt() |> tab_header(title = "Basin-scale climate indicators, current year") |> sub_missing(missing_text = "—")
```
### 5.5 Demographic composition of take
```{r}
ctx_demo |> gt() |> tab_header(title = "Size-class composition, mortality and population context") |> sub_missing(missing_text = "—")
```

## 6. ITS running-sum tracker

```{r its}
its_track <- per_species |>
  inner_join(its_ref |> filter(!is.na(annual_its)), by = "species_code") |>
  rowwise() |>
  mutate(wyrs = list(seq(window_start_year, length.out = its_window_years)),
         cov = list(if (sector == "DSLL") effort$coverage_rate[match(unlist(wyrs), effort$year)] else rep(1, its_window_years)),
         obs = list(ann$interactions[ann$species_code == species_code][match(unlist(wyrs), ann$year[ann$species_code == species_code])]),
         years_elapsed = sum(!is.na(unlist(obs))),
         cumulative = sum(unlist(obs) / unlist(cov), na.rm = TRUE),
         expansion = 1 / mean(unlist(cov), na.rm = TRUE),
         window_limit = annual_its * its_window_years,
         yrs_left = its_window_years - years_elapsed,
         p_exceed = if (yrs_left > 0) 1 - pnbinom((window_limit - cumulative) / expansion, size = prior_shape * yrs_left,
                                                  prob = prior_rate / (prior_rate + 1)) else as.numeric(cumulative > window_limit),
         alert = p_exceed > 0.25) |>
  ungroup() |>
  select(Species = common, Authority = authority, `Annual ITS` = annual_its, `Window (yrs)` = its_window_years,
         `Years elapsed` = years_elapsed, `Cumulative expanded take` = cumulative, `Window limit` = window_limit,
         `P(exceed before close)` = p_exceed, Alert = alert)
its_track |> gt() |> fmt_number(columns = 6, decimals = 0) |> fmt_number(columns = 8, decimals = 3) |>
  tab_source_note("ITS values are ILLUSTRATIVE placeholders; replace with the governing BiOp values. Alert threshold 0.25 follows the working-group trigger spec.")
```

## 7. Data aggregation and key findings {#sec-aggregation}

This chapter aggregates Tables 2, 2b and 2c, the Bayesian check, the ITS tracker and the working-group
context tables into the findings that lead the SAFE narrative. It is computed, not written, so it changes
with the data. (When assembling the report, move this chapter above the tables with `{{< include >}}` or
by reordering the child documents.)

```{r aggregation, results="asis"}
A <- per_species |> left_join(table2b |> select(species_code, rate_robust_anomaly, rate_percentile, annual_pct_change,
                                              trend_model_flag, centroid_shift_km, pct_outside_hist90), by = "species_code")
flagged  <- filter(A, flag == "High"); lowf <- filter(A, flag == "Low")
gates_of <- function(r) paste(c("robust","percentile","bayes")[c(r$gate_robust, r$gate_percentile, r$gate_bayes)], collapse = ", ")
beyond   <- filter(A, current > pred_q95 | current < pred_q05)
bigshift <- filter(A, centroid_shift_km > 250) |> arrange(desc(centroid_shift_km))
outside  <- filter(A, pct_outside_hist90 > 10)
ocean_reg <- table2c |> group_by(variable) |>
  summarise(mean_pct = mean(interaction_percentile), hi = sum(interaction_percentile >= 75), lo = sum(interaction_percentile <= 25),
            sel_x = sum(selection_percentile >= 90 | selection_percentile <= 10), n = n(), .groups = "drop") |>
  mutate(direction = case_when(hi > lo ~ "high", lo > hi ~ "low", TRUE ~ "mixed"))
fallb <- filter(A, grepl("fallback|error", trend_model_flag, ignore.case = TRUE))
eff_top <- ctx_spatial |> slice_max(abs(shift_pct_points), n = 1)
its_hi  <- its_track |> filter(`Cumulative expanded take` / `Window limit` >= 0.75)

findings <- c(
  if (nrow(flagged)) sprintf("%d of %d monitored species %s flagged High in %d: %s.", nrow(flagged), nrow(A),
        ifelse(nrow(flagged) == 1, "is", "are"), CUR,
        paste(sprintf("%s (%d vs median %d; gates: %s)", flagged$common, flagged$current, as.integer(flagged$median),
                      sapply(seq_len(nrow(flagged)), \(i) gates_of(flagged[i, ]))), collapse = "; "))
  else sprintf("No species is flagged High in %d; all counts fall within their historical distributions.", CUR),
  if (nrow(lowf)) sprintf("%s flagged Low (robust anomaly <= -2).", paste(lowf$common, collapse = ", ")),
  sprintf("%d of %d species are above their historical median and %d show an increasing long-run slope; the rate anomaly agrees in sign with the count anomaly for %d of %d, so most of the change is not an artefact of effort.",
          sum(A$current > A$median), nrow(A), sum(A$trend == "Increasing"),
          sum(sign(A$robust_anomaly) == sign(A$rate_robust_anomaly)), nrow(A)),
  if (nrow(beyond)) sprintf("The current count falls outside the Gamma-Poisson posterior predictive 90%% interval for %s.",
        paste(sprintf("%s (%s)", beyond$common, ifelse(beyond$current > beyond$pred_q95, "above", "below")), collapse = ", "))
  else "No current count falls outside its Gamma-Poisson posterior predictive 90% interval.",
  if (nrow(bigshift)) sprintf("Interaction centroids moved more than 250 km for %s%s.",
        paste(sprintf("%s (%.0f km)", bigshift$common, bigshift$centroid_shift_km), collapse = ", "),
        if (nrow(outside)) sprintf("; %s fell outside the historical 90%% area",
              paste(sprintf("%.1f%% of %s interactions", outside$pct_outside_hist90, tolower(outside$common)), collapse = " and ")) else "")
  else "Interaction centroids stayed within 250 km of their historical positions for every species.",
  { oc <- filter(ocean_reg, hi >= ceiling(n * .6) | lo >= ceiling(n * .6))
    if (nrow(oc)) sprintf("A coherent ocean regime is visible at interaction sets: %s.",
        paste(sprintf("%s %s for %d of %d species (mean %.0fth percentile)", oc$variable, oc$direction,
                      ifelse(oc$direction == "high", oc$hi, oc$lo), oc$n, oc$mean_pct), collapse = "; ")) },
  { sx <- filter(ocean_reg, sel_x > 0)
    if (nrow(sx)) sprintf("Fleet selection anomalies are at historical extremes for %s, indicating the fleet fished a different habitat envelope than usual.",
        paste(sprintf("%s (%d species)", sx$variable, sx$sel_x), collapse = ", ")) },
  if (nrow(eff_top)) sprintf("Fleet effort share changed most in the %s stratum (%+.1f points; %s).", eff_top$stratum, eff_top$shift_pct_points, eff_top$operational_driver),
  if (nrow(its_track)) { al <- filter(its_track, Alert)
    if (nrow(al)) sprintf("ITS running-sum alert for %s.", paste(al$Species, collapse = ", "))
    else if (nrow(its_hi)) sprintf("No ITS running-sum alert; %s.", paste(sprintf("%s stands at %.0f%% of its window limit after %d of %d years",
              its_hi$Species, 100 * its_hi$`Cumulative expanded take` / its_hi$`Window limit`, its_hi$`Years elapsed`, its_hi$`Window (yrs)`), collapse = ", "))
    else "No ITS running-sum alert; all tracked species are below 75% of their window limits." },
  if (nrow(fallb)) sprintf("Negative-binomial trend fitting fell back to Poisson for %s; those annual-percent-change values carry wider uncertainty.", paste(fallb$common, collapse = ", "))
)
cat("### Key findings\n\n", paste0("* ", findings, "\n"), sep = "")
```

```{r aggregation-table}
A |> transmute(Species = common, Flag = flag, `Gates fired` = sapply(seq_len(n()), \(i) gates_of(A[i, ])),
               `Count (percentile)` = sprintf("%d (P%.0f)", current, percentile),
               `Rate anomaly` = sprintf("%+.2f (P%.0f)", rate_robust_anomaly, rate_percentile),
               Trend = sprintf("%+.1f%%/yr", annual_pct_change), `Centroid shift` = sprintf("%.0f km", centroid_shift_km),
               `Predictive 90%` = sprintf("%d–%d", pred_q05, pred_q95)) |>
  gt() |> tab_header(title = md(sprintf("**Cross-species aggregation of anomaly evidence, %d**", CUR))) |>
  data_color(columns = Flag, fn = function(x) case_when(x == "High" ~ "#F18F01", x == "Low" ~ "#1C7293", TRUE ~ "#F4F6F8"))
ocean_reg |> transmute(Variable = variable, `Mean percentile` = round(mean_pct), `Species high (≥75th)` = sprintf("%d of %d", hi, n),
                       `Species low (≤25th)` = sprintf("%d of %d", lo, n), `Extreme selection anomaly` = sel_x,
                       Reading = ifelse(direction == "mixed", "no coherent signal", paste(direction, "across the fishery"))) |>
  gt() |> tab_header(title = "Ocean regime at interaction sets")
```

## 8. Draft causal narratives {#sec-narratives}

The narrative for each species is drafted from the tables above. With `params$use_llm = TRUE` the
aggregated table rows for each species are sent to a large-language model to produce a *first draft*
that an analyst then edits; with `use_llm = FALSE` a rule-based template is used.

::: {.callout-important}
## AI use governance
Only aggregated summary statistics (the rows of Tables 2, 2b, 2c) are sent to the model — never PIROP observer
records. NOAA analysts must route this call through an agency-approved, FedRAMP-authorized endpoint
(set `params$llm_endpoint`, e.g. a Vertex AI project) rather than a public consumer service, and every draft
carries a provenance line so it cannot be mistaken for reviewed text.
:::

```{r gemini-narratives}
#| echo: true
#| code-fold: false

# Build a compact, aggregated evidence packet per species -----------------------
evidence_packet <- function(code) {
  r2  <- per_species |> filter(species_code == code)
  r2b <- table2b     |> filter(species_code == code)
  oc  <- table2c     |> filter(species_code == code)
  list(
    species = r2$common, year = CUR, flag = r2$flag,
    interactions = list(current = r2$current, median = r2$median, p10 = r2$p10, p90 = r2$p90,
                        percentile = r2$percentile, robust_anomaly = round(r2$robust_anomaly, 2),
                        bayes_tail_prob = round(r2$p_tail, 3), posterior_pred_90 = c(r2$pred_q05, r2$pred_q95)),
    rate = list(current = r2b$rate_cur, percentile = r2b$rate_percentile, robust_anomaly = round(r2b$rate_robust_anomaly, 2),
                annual_pct_change = round(r2b$annual_pct_change, 1), trend_model = r2b$trend_model, trend_years = r2b$tw),
    spatial = list(centroid_shift_km = round(r2b$centroid_shift_km), pct_outside_hist90 = r2b$pct_outside_hist90),
    ocean = oc |> transmute(variable, interaction_percentile = round(interaction_percentile),
                            selection_percentile = round(selection_percentile), absolute_anomaly = signif(absolute_anomaly, 3)),
    context = list(spatial_effort = ctx_spatial, gear = ctx_gear, behaviour = ctx_behavior,
                   climate = ctx_climate, demographics = filter(ctx_demo, species_code == code))
  )
}

system_prompt <- paste(
  "You are drafting the protected-species interactions narrative for the Western Pacific Regional Fishery",
  "Management Council annual SAFE report (Pelagic FEP section 3.3.2). Write in plain, factual, third-person",
  "scientific prose for a Council audience. Use ONLY the numbers in the evidence packet; do not invent values.",
  "Structure: (1) one sentence stating the flag and the count vs. historical distribution; (2) whether the rate,",
  "trend and Bayesian check agree; (3) spatial and ocean context and which causal-diagnostic branch the evidence",
  "supports (fishery change, expected from trend/variability, ocean-driven distribution or catchability shift,",
  "or cause unclear); (4) what SME review is needed. Cite the causal-diagnostic branch explicitly. 120-180 words.",
  "Do not use hedging filler. Do not mention that you are an AI.")

draft_with_gemini <- function(packet) {
  key <- Sys.getenv("GEMINI_API_KEY")
  stopifnot("GEMINI_API_KEY not set" = nzchar(key))
  body <- list(
    system_instruction = list(parts = list(list(text = system_prompt))),
    contents = list(list(role = "user", parts = list(list(text = paste0(
      "Evidence packet (JSON):\n", jsonlite::toJSON(packet, auto_unbox = TRUE, pretty = TRUE, digits = 4)))))),
    generationConfig = list(temperature = 0.2, maxOutputTokens = 600))
  resp <- httr2::request(sprintf("%s/models/%s:generateContent", params$llm_endpoint, params$llm_model)) |>
    httr2::req_headers(`x-goog-api-key` = key, `Content-Type` = "application/json") |>
    httr2::req_body_json(body) |>
    httr2::req_retry(max_tries = 3) |>
    httr2::req_perform() |>
    httr2::resp_body_json()
  txt <- resp$candidates[[1]]$content$parts[[1]]$text
  paste0(trimws(txt), "\n\n*Draft generated by ", params$llm_model, " on ", Sys.Date(),
         " from aggregated pipeline outputs; requires analyst review before use.*")
}

narratives <- map(species$species_code, function(code) {
  pk <- evidence_packet(code)
  txt <- if (isTRUE(params$use_llm)) tryCatch(draft_with_gemini(pk), error = function(e) {
             paste0("[LLM call failed: ", conditionMessage(e), "] ",
                    template_narrative(filter(per_species, species_code == code),
                                       filter(table2b, species_code == code), filter(table2c, species_code == code)))
           }) else template_narrative(filter(per_species, species_code == code),
                                      filter(table2b, species_code == code), filter(table2c, species_code == code))
  list(species_code = code, species = pk$species, flag = pk$flag, narrative = txt,
       source = if (isTRUE(params$use_llm)) params$llm_model else "rule-based template")
})
```

```{r print-narratives, results="asis"}
for (n in narratives) cat(sprintf("\n### %s — %s\n\n%s\n", n$species, n$flag, n$narrative))
```

## 9. Export for the decision-support application

```{r export}
out <- list(
  meta = list(generated = as.character(Sys.time()), current_year = CUR, alpha = ALPHA,
              mode = if (params$use_mock) "test" else "lotus",
              decision_rule = sprintf("OR gate: robust >= %s | percentile >= %s | bayes p < %s",
                                      params$robust_threshold, params$percentile_threshold, ALPHA)),
  table2_interactions = per_species |> select(species_code, species = common, current, mean, median, p10, p90, percentile,
                                              robust_anomaly, n_years, trend_slope, trend, flag),
  table2b_rate_spatial = table2b |> select(species_code, species = common, current_rate = rate_cur, hist_median_rate, hist_p90_rate,
                                           rate_percentile, rate_robust_anomaly, annual_pct_change, trend_model, trend_model_flag,
                                           trend_years = tw, centroid_shift_km, pct_outside_hist90),
  table2c_ocean = table2c,
  bayes = per_species |> select(species_code, species = common, prior_shape, prior_rate, posterior_pred_q05 = pred_q05,
                                posterior_pred_median = pred_median, posterior_pred_q95 = pred_q95, p_tail) |> mutate(alpha = ALPHA, flagged = p_tail < ALPHA),
  its = its_track,
  aggregation = list(findings = findings, ocean_regime = ocean_reg),
  narratives = narratives,
  series = split(ann, ann$species_code) |> map(~ list(years = .x$year, interactions = .x$interactions)),
  effort = effort,
  context = list(spatial_effort = ctx_spatial, gear = ctx_gear, fisher_behavior = ctx_behavior,
                 climate_indicators = ctx_climate, demographics = ctx_demo)
)
jsonlite::write_json(out, params$export_json, auto_unbox = TRUE, pretty = TRUE, digits = 6, na = "null")
cat("Wrote", params$export_json, "— load this file in the decision-support app (Data Pipeline tab).")
```

## Appendix — model notes

* **Robust anomaly** uses the median and MAD (×1.4826) so a single extreme year cannot inflate the baseline.
* **Trend** is a log-linear Poisson GLM with observed sets as an offset; a negative-binomial model replaces it
  when the Pearson dispersion exceeds 1.5. When `glm.nb` fails to converge the Poisson fit is retained and the
  table flags it, matching the "NB fitting error; Poisson fallback" convention in the reference workbook.
* **Bayesian check** is a Gamma–Poisson conjugate model. The prior is method-of-moments from the historical years
  (variance-matched when overdispersed). The tail probability P(X ≥ current) is evaluated with
  `1 - pbeta(p, shape, current)`, the negative-binomial–incomplete-beta identity, which is exact for non-integer shape
  (spreadsheet `NEGBINOM.DIST` silently truncates it).
* **OR gate**: any single gate firing yields *High*. The gate thresholds are parameters so they can be
  re-elicited with the SDM swing-weighting tool in the decision-support app.
* **Placeholders needing SME validation before Council use**: ITS values, gear/bait multipliers, demographic
  size-class shares, and the ocean selection-anomaly interpretation.
